---
title: "Calculating distribution of atac-seq counts per peak"
output: html_notebook
---

Having used the merged peak file and featureCounts to get a matrix of the fragments that overlap each peak (taking all peaks and samples into account, that is, not caring if a peak is not present in a sample), I want to get the distribution of fragments (paired end reads) per peak. The y axis will be the frequency, and the x axis will be numerical and discrete, and ranging to 0 to the maximum counts.

```{r setup,include=FALSE}

currentDate <- Sys.Date() # to save date in name of output files


counts <- read.table(file="/Users/Marta/Documents/WTCHG/DPhil/Data/Regulation/atac-seq/plotCoverage/counts_all_samples_peaks.txt",header = T) # bed file with start and end of peaks

samples=c("A","B","C","D")
samples_2=rep(samples,each=6)
stage=as.character(c(1:6))
stage_2=rep(stage,4)

colnames(counts)[7:ncol(counts)]=paste(samples_2,stage_2,sep="")  # rename columns for samples and stages

```

```{r summarise, include=FALSE}
mean_peak_length=mean(counts$Length)  # calculate mean peak length

each_sample_counts_mean=colMeans(counts[7:ncol(counts)]) # calculate mean counts per sample
appended_columns=unlist(counts[7:ncol(counts)],use.names = F)
all_sample_counts_mean=mean(appended_columns)   # calculate mean counts for all samples
# also, for each peak:
each_peak_counts_mean=rowMeans(counts[7:ncol(counts)])
mean(each_peak_counts_mean)  # gives same result as above

all_sample_counts_variance=var(appended_columns) # calculate variance for counts for all samples

# mode
getmode <- function(v) {
   uniqv <- unique(v)
   uniqv[which.max(tabulate(match(v, uniqv)))]
}

mode= getmode(appended_columns)

```
The mean peak length is `r round(mean_peak_length,digits=2)`.
The mean counts per sample are `r round(each_sample_counts_mean,digits=2)`.
The mean counts for all samples is `r round(all_sample_counts_mean,digits=2)`.
The variance of counts for all samples is `r round(all_sample_counts_variance,digits=2)`.

```{r plot,echo=FALSE,message=FALSE,warning=FALSE}

library(ggplot2)
library(grid)
library(gridExtra)

unique_values_counts=aggregate(data.frame(count = appended_columns), list(value = appended_columns), length)

colnames(unique_values_counts)=c("counts_in_peak","number")

sum(unique_values_counts$number)  # total times a counts_in_peak values is repeated 
# 5394696

# frequency for each unique value of counts per peak:

unique_values_counts$frequency=unique_values_counts$number/sum(unique_values_counts$number)

sum(unique_values_counts$frequency) # sum of all frequencies gives 1 

p1 <- ggplot(unique_values_counts, aes(x=counts_in_peak, y=frequency))
p1 <- p1 + geom_point() + scale_x_continuous(limits = c(0, 100))
plot(p1)

ggsave(paste("/Users/Marta/Documents/WTCHG/DPhil/Plots/atac-seq/distribution_peak_counts_frequency_first_100",currentDate,".jpg",
             sep=""),p1,width=8,height=8,units="in",dpi=300)

p2 <- ggplot(unique_values_counts, aes(x=counts_in_peak, y=number))
p2 <-p2 + geom_point() + scale_x_continuous(limits = c(0, 100))
plot(p2)

p3 <- ggplot(unique_values_counts, aes(x=counts_in_peak, y=frequency))
p3 <-p3 + geom_point() + scale_x_continuous(limits=c(101,23150)) + scale_y_continuous(limits = c(0, 0.00060))
plot(p3)

p4 <- ggplot(unique_values_counts, aes(x=counts_in_peak, y=number))
p4 <-p4 + geom_point() + scale_x_continuous(limits=c(101,23150)) + scale_y_continuous(limits = c(0, 3400))
plot(p4)

distribution_peak_counts=grid.arrange(p1, p2, p3, p4, ncol=4, nrow = 2, layout_matrix = rbind(c(1,1,2,2), c(3,3,4,4)),widths = c(2,1,1,2), heights = c(3,3))
ggsave(paste("/Users/Marta/Documents/WTCHG/DPhil/Plots/atac-seq/distribution_peak_counts",currentDate,".jpg",sep=""),distribution_peak_counts,width=8,height=8,units="in",dpi=300)

```

```{r save,echo=FALSE,message=FALSE,warning=FALSE}


#read in table of names

names <- read.table(file="/Users/Marta/Documents/WTCHG/DPhil/Data/Regulation/atac-seq/names_samples.txt",header = T)

names$letter_number <- paste(names$letter,names$number,sep="")

new_counts=counts[7:ncol(counts)]

colnames(new_counts)=paste(names$stage,names$sample,sep="-")


write.table(file="/Users/Marta/Documents/WTCHG/DPhil/Data/Regulation/atac-seq/counts_all_samples_peaks_renamed.txt",new_counts,col.names = T,row.names = F,sep="\t")

```



