---
title: "Exploratory QC"
output: html_notebook
---

Checking using CPM>1 in all three replicates for at least one stage. Let's see how many I've got left after filtering.

```{r setup, include=FALSE, warning=FALSE}
library(edgeR)
library(knitr)   # for kable tables

currentDate <- Sys.Date() # to save date in name of output files

load(file="/Users/Marta/Documents/WTCHG/DPhil/Data/Regulation/atac-seq/session_objects/dge_atac-seq.xz",verbose=TRUE)  #loading the dge object for conservative counts



```

Conservative counts initially has `r nrow(dge)` peaks, and looks like this: 


```{r table, warning=FALSE}


kable(dge$counts[1:2,1:6])


```

Now I want to keep in my count object peaks with at least 1 cpm in at least one stage.

```{r first_trim, warning=FALSE}

old_peaks=dim(dge)[1]# save how many peaks there are initially
keep <- rowSums(cpm(dge,normalized.lib.sizes = TRUE)>1) >= 3   # getting peaks with at least 1 cpm in at least one stage (3 rep per stage).
dge <- dge[keep, , keep.lib.sizes=FALSE]  # adjusting the library size

dge <- calcNormFactors(dge)# recalculating normalization factors after altering library sizes
# if counts are distributed roughly equally across all samples, it shouldn0t change much

percent=(dim(dge)[1]/old_peaks)*100 # percentage from original pekas that we conserve



```

Conservative counts now has `r nrow(dge)` peaks (`r percent`% of original amount). Checking that I have removed all peaks in which at least one stage doesn't have 1 CPM for ALL 3 samples (I wasn't really checking that the cpm>1 were in the same stage).

```{r second_trim, warning=FALSE}

# go through the table 3 by 3
# if there are more than 2 values in three columns above 0, go to next row.
# if not, go to next 3
# if gets to end of loop and not found the values, get that row. After loop, subset matrix excluding those rows

 cpm_data=cpm(dge, normalized.lib.sizes=TRUE)

trim_by_stage=function(dge){
  data=cpm(dge, normalized.lib.sizes=TRUE)
   data[data<1]=0
  data[data>1]=1 # convert to table where 0=0 and 1 equals any value above n
  c=c() # initiate vector of numbers, to save rows to take out of matrix

  for (r in 1:nrow(data)){
   for (i in seq(1,24,3)){   # loop through stages 3 by 3 (to jump over samples of same stage)

     if (sum(data[r,c(i,i+1,i+2)]) > 2) {

        break
      }
      if (i==22){

        c=c(c,r)  # if it gets to the end and there are no valid values, append to vector
      }


   }

  }

rm(data)
return(c)

}


c=trim_by_stage(dge)
fail=100*(length(c)/nrow(dge$counts))
genes_fail=length(c)
ditch=dge$counts[c,] # table to throw away

keep=!(rownames(dge$counts) %in% rownames(ditch))

dge <- dge[keep, , keep.lib.sizes=FALSE] # counts now has 17892 genes

rm(c,keep,ditch)


```

After trimming, peak counts has `r nrow(dge)` peaks. It has lost `r genes_fail` genes (`r fail`% from previous total). We keep `r (nrow(dge)/old_peaks)*100`% of the original peaks.

Plot the distribution of CPM trimming and percentage of peaks that remain from the original amount (from CPM 1 to 7 or so).

```{r full trimming function, warning=FALSE}
load(file="/Users/Marta/Documents/WTCHG/DPhil/Data/Regulation/atac-seq/session_objects/dge_atac-seq.xz",verbose=TRUE)  #loading the dge object for conservative counts

old_peaks=dim(dge)[1]# save how many peaks there are initially

trim_by_stage_and_cpm=function(dge,CPM=1,old_peaks=dim(dge)[1]){
keep <- rowSums(cpm(dge,normalized.lib.sizes = TRUE)>CPM) >= 3   # getting peaks with at least n cpm in at least one stage (3 rep per stage).
dge <- dge[keep, , keep.lib.sizes=FALSE]  # adjusting the library size

dge <- calcNormFactors(dge)# recalculating normalization factors after altering library sizes
# if counts are distributed roughly equally across all samples, it shouldn0t change much

# go through the table 3 by 3
# if there are more than 2 values in three columns above 0, go to next row.
# if not, go to next 3
# if gets to end of loop and not found the values, get that row. After loop, subset matrix excluding those rows

 cpm_data=cpm(dge, normalized.lib.sizes=TRUE)

trim_by_stage=function(dge){
  data=cpm(dge, normalized.lib.sizes=TRUE)
  data[data<CPM]=0
  data[data>CPM]=1 # convert to table where 0=0 and 1 equals any value above n
  
  c=c() # initiate vector of numbers, to save rows to take out of matrix

  for (r in 1:nrow(data)){
   for (i in seq(1,24,3)){   # loop through stages 3 by 3 (to jump over samples of same stage)

     if (sum(data[r,c(i,i+1,i+2)]) > 2) {

        break
      }
      if (i==22){

        c=c(c,r)  # if it gets to the end and there are no valid values, append to vector
      }


   }

  }

rm(data)
return(c)

}

c=trim_by_stage(dge)  # calling previous function

fail=100*(length(c)/nrow(dge$counts))
genes_fail=length(c)
ditch=dge$counts[c,] # table to throw away

keep=!(rownames(dge$counts) %in% rownames(ditch))

dge <- dge[keep, , keep.lib.sizes=FALSE] # counts now has 17892 genes

rm(c,keep,ditch)

peaks_remain=((nrow(dge)/old_peaks)*100)

return(peaks_remain)
}

# call function for all CPMs
peaks_remain=c()
for(i in 1:7){
  peaks_remain[i]=trim_by_stage_and_cpm(dge,CPM=i,old_peaks=dim(dge)[1])
}


df=data.frame(CPM = c(1:7),peaks_remain=peaks_remain)

library(ggplot2)

p <- ggplot(df,aes(y=peaks_remain, x=CPM)) +
  geom_line() + xlab("CPM cutoff") + ylab("% of peaks that remain") + 
  scale_x_continuous(limits=c(1,7),breaks=c(1:7)) + ylim(c(0,100)) 
  
ggsave(paste("/Users/Marta/Documents/WTCHG/DPhil/Plots/atac-seq/CPM_cutoff_peaks_remaining",currentDate,".jpg",
             sep=""),p,width=4,height=4,units="in",dpi=300)


```


